Probabilistic Inference of Basic Stellar Parameters:\\ Application to Flickering Stars \altaffilmark{\titledag}

### 2015-12-02 version1.0/manuscript.tex
now with Dave's comments, about to submit

%% Authors
\author{
	{\bf	Ruth~Angus\altaffilmark{1,2},
		David.~M.~Kipping\altaffilmark{2,3},
	}
}

@@

\altaffiltext{2}{Harvard-Smithsonian Center for Astrophysics,
		Cambridge, MA 02138, USA; email: dkipping@cfa.harvard.edu}

\altaffiltext{3}{Department of Astronomy, Columbia University, 550 W 120th St.,
                 New York, NY 10027, US}

@@

The demand for probilistic stellar parameters is not only motivated by the fact
that probability distributions are far more representative of our `beliefs'
about astrophysical parameters, it also has a practical purpose.
When using data published in the astronomical literature to, for example, infer
relationships between observed parameters, that inference can be performed as
the final stage in a hierarchical treatment
\citep[see, e.g.][]{foreman-mackey:2014}.
This is useful if, for example, one wants to account for multi-dimensional
uncertainties by marginalizing over the `true' parameter values upon which the
noisy observations are conditioned.
This point is discussed further in \textsection\ref{sec:HBM}.

@@

Whether one calibrates flicker to \logg\ or \rhostar, there are several aspects
of the problem which are attractive for our purposes of a simple demonstration
of probabilistic inference of stellar parameters.
Firstly, in log-log space the relationship is very simple, appearing to be
linear \citep{kipping:2014}.
Secondly, there is a sufficiently large number of points in the sample (439
stars) to constrain a population-based model.
Thirdly, there is significant excess scatter around the best-fitting relation
implying that a deterministic model is inadequate.
This is not surprising given that granulation is a complex and messy process
for which one should not expect any parametric model to provide a perfect
description.
Fourthly, the physical processes that produce surface granulation, of which
flicker is an observational tracer, may be more or less noisy for different
types of stars.
We will test whether flicker has greater predictive power in certain regions
of parameter space; i.e. is flicker significantly more informative for
subgiants than for dwarfs?
For these reasons, we identify the calibration of
flicker to \logg\ and \rhostar\ as a well-posed problem to first demonstrate
probabilistic inference in the arena of stellar characterization.

@@

We also performed this inference using an alternative, analytic form of the
marginal likelihood, which accounts for 2-D uncertainties but does not allow
the intrinsic scatter to be a function of the dependent or independent
variables.
For the relation between flicker and \rhostar, this likelihood function can be
written as
\begin{eqnarray}
	& p(\rho_*| F_8, \alpha, \beta, \sigma_{\rho}) =  \\ \nonumber
						      & \exp \left[-\frac{1}{2}
		\sum_{n=1}^N \frac{[F_{8n}-(\alpha + \beta \rho_{*n})]^2}
	{\left[\beta \sigma_{F8, n}^2 + (\sigma_{\rho *, n}
	+ \sigma_{\rho})^2\right]}\right]
	\\ \nonumber
	& + \beta^2 \sigma_{F8, n}^2 + (\sigma_{\rho *, n} + \sigma_{\rho})^2,
\end{eqnarray}
\label{eq:likelihood}
and similarly for \logg.
The parameter values found using this simple likelihood function are included
in table \ref{tab:results}.
% The analytic likelihood function, equation \ref{eq:likelihood} is suitable for
% cases where the intrinsic scatter is not a function of $x$ or $y$.
Figures \ref{fig:rhostar} and \ref{fig:logg} show the data with the best-fit
models.
The shaded regions show the 1 and 2$\sigma$ confidence interval which are
representative of the intrinsic scatter in the relations.
% Given a measurement of flicker, the predictive probability distribution of
% \logg and \rhostar will be a Gaussian distribution

@@

\begin{table}
\caption{Highest-likelihood parameter values with 16th and 84th
	percentile uncertainties.}
\begin{tabular}{lccc}
\hline\hline
Parameter & highest-likelihood value & simple model result \\
    \hline
$\alpha$ &    2.358$_{-0.04}^{+0.08}$ &       2.191$_{-0.05}^{+0.05}$  \\
$\beta$ &    -1.79$_{-0.03}^{+0.04}$ &        -1.724$_{-0.03}^{+0.03}$ \\
$\sigma_{rho}$ &    0.015$_{-0.04}^{+0.08}$ &   0.085$_{-0.05}^{+0.05}$  \\
$\gamma$ &    -0.0$_{-0.04}^{+0.08}$ & \\
$\delta$ &    5.535$_{-0.07}^{+0.07}$ &      5.739$_{-0.03}^{+0.03}$  \\
$\epsilon$ &    -1.005$_{-0.04}^{+0.03}$ &   -1.095$_{-0.02}^{+0.01}$ \\
$\sigma_g$ &    0.168$_{-0.07}^{+0.07}$ &    0.015$_{-0.03}^{+0.03}$  \\
$\zeta$ &    -0.014$_{-0.07}^{+0.07}$ & \\
    \hline
\label{tab:results}
\end{tabular}
\end{table}

@@

We have recalibrated the relation between short timescale brightness
fluctuations in the {\it Kepler} light curves of stars (flicker) with both
stellar density and surface gravity, whilst including parameters to describe
the intrinsic scatter in these relationships.
We included both an additive variance term {\it and} a term that allows for
some abscissa-dependent variance, presented in table \ref{tab:table}.
The additive terms, $\sigma_\rho$ and $ \sigma_g$ are both non-zero,
suggesting that there {\it is} an additional source of scatter in the
relations, not accounted for by the observational uncertainties alone.
This is either caused by intrinsic scatter in the physical relationship
between flicker and density and \logg, produced by some physical process that
is not accounted for in the model, or by an underestimation of the
observational uncertainties.
Interestingly, the abscissa-dependent variances, $\gamma$ and $\zeta$ are both
less than zero.
This indicates that the intrinsic scatter either decreases with increasing
flicker and decreasing stellar density and surface gravity, or that the
uncertainty underestimation is more severe for the giant stars.
Regardless of the cause of the additional scatter needed to describe the
relations between these parameters, the conclusions of this study are the same.
Firstly, naively modelling the relationship between flicker and surface
gravity or stellar density, or any other stochastic relationship
deterministically, will lead to an underprediction of the uncertainties in
dependent variable.
Secondly, we find no evidence for a increased scatter at higher surface
gravities and densities, i.e. the relations between flicker, surface gravity
and stellar density are equally valid for dwarfs and giants alike.

This is a simple `fitting a line to data' exercise, however it continues the
discussion of probabilistic modelling that is an active topic within the
fields of exoplanet and stellar astronomy.
We use Hierarchical Bayesian Modelling (HBM) to constrain the intrinsic
scatter in the relationship between flicker, surface gravity and density.
We also include the effects of the non-negligable two-dimensional observational
uncertainties by marginalizing over latent variables using importance sampling.
% Although none of these methods, nor the data used here are new taken alone,
% we use the example of flicker to demonstrate the ideas and the importance of
% probabilistic, hierarchical inference.
Relationships between astronomical parameters are almost always
non-deterministic; an element of stochasticity effects the physical parameters
of stars so one can never perfectly predict $y$ given an observation of $x$.
We advocate a probabilistic approach in both the `fitting the model to data'
step, {\it and} when using an empirically calibrated model to predict
parameter values.
The fitting stage benefits because if the relationships between parameters are
falsely assumed to be deterministic, they will be skewed by data points with
unrepresentative uncertainties.
The prediction stage benefits from the stochastic treatment both because a
probability distribution is in many ways more representative of an observation
than a point estimate, and because posterior PDF samples can be used in
subsequent studies (provided the prior used during the fitting process is
described).

>>>>

%% Authors
\author{
	{\bf	Ruth~Angus\altaffilmark{1},
		David.~M.~Kipping\altaffilmark{2},
	}
}

@@

\altaffiltext{2}{Department of Astronomy, Columbia University, 550 W 120th St.,
                 New York, NY 10027, US}

@@

The demand for probabilistic stellar parameters is not only motivated by the fact
that probability distributions are far more representative of our `beliefs'
about astrophysical parameters, it also has a practical purpose.
When using data published in the astronomical literature to, for example, infer
relationships between observed parameters, that inference can be performed as
the final stage in a hierarchical treatment
\citep[see, e.g.][]{foreman-mackey:2014}.
This is useful if, for example, one wants to account for multi-dimensional
uncertainties by marginalizing over the `true' parameter values upon which the
noisy observations are conditioned.
This point is discussed further in \textsection\ref{sec:HBM}.

@@

Whether one calibrates flicker to \logg\ or \rhostar, there are several aspects
of the problem which are attractive for our purposes of a simple demonstration
of probabilistic inference of stellar parameters.
Firstly, in log-log space the relationship is very simple, appearing to be
linear \citep{kipping:2014}.
Secondly, there is a sufficiently large number of points in the sample (439
stars) to constrain a population-based model.
Thirdly, there is significant excess scatter around the best-fitting relation
implying that a deterministic model is inadequate.
This is not surprising given that granulation is a complex and messy process
for which one should not expect any parametric model to provide a perfect
description.
Finally, the physical processes that produce surface granulation, of which
flicker is an observational tracer, may be more or less noisy for different
types of stars.
We will test whether flicker has greater predictive power in certain regions
of parameter space; i.e. is flicker significantly more informative for
subgiants than for dwarfs?
For these reasons, we identify the calibration of
flicker to \logg\ and \rhostar\ as a well-posed problem to first demonstrate
probabilistic inference in the arena of stellar characterization.

@@

We also performed this inference using an alternative, analytic form of the
marginal likelihood, which accounts for 2-D uncertainties but does not allow
the intrinsic scatter to be a function of the dependent or independent
variables.
For the relation between flicker and \rhostar, this likelihood function can be
written as
\begin{eqnarray}
	& p(\rho_*| F_8, \alpha, \beta, \sigma_{\rho}) =  \\ \nonumber
						      & \exp \left[-\frac{1}{2}
		\sum_{n=1}^N \frac{[F_{8n}-(\alpha + \beta \rho_{*n})]^2}
	{\left[\beta \sigma_{F8, n}^2 + (\sigma_{\rho *, n}
	+ \sigma_{\rho})^2\right]}\right]
	\\ \nonumber
	& + \beta^2 \sigma_{F8, n}^2 + (\sigma_{\rho *, n} + \sigma_{\rho})^2,
\end{eqnarray}
\label{eq:likelihood}
and similarly for \logg.
The parameter values found using this simple likelihood function, with a
constant dispersion are included in table \ref{tab:results}.
% The analytic likelihood function, equation \ref{eq:likelihood} is suitable for
% cases where the intrinsic scatter is not a function of $x$ or $y$.
Figures \ref{fig:rhostar} and \ref{fig:logg} show the data with the best-fit
models.
The shaded regions show the 1 and 2$\sigma$ confidence interval which are
representative of the intrinsic scatter in the relations.
% Given a measurement of flicker, the predictive probability distribution of
% \logg and \rhostar will be a Gaussian distribution

@@

\begin{table}
\caption{Highest-likelihood parameter values with 16th and 84th
	percentile uncertainties.}
\begin{tabular}{lccc}
\hline\hline
Parameter & Variable dispersion model & Constant dispersion model \\
    \hline
$\alpha$ &    2.358$_{-0.04}^{+0.08}$ &       2.191$_{-0.05}^{+0.05}$  \\
$\beta$ &    -1.79$_{-0.03}^{+0.04}$ &        -1.724$_{-0.03}^{+0.03}$ \\
$\sigma_{rho}$ &    0.015$_{-0.04}^{+0.08}$ &   0.085$_{-0.05}^{+0.05}$  \\
$\gamma$ &    -0.0$_{-0.04}^{+0.08}$ & \\
\hline
$\delta$ &    5.535$_{-0.07}^{+0.07}$ &      5.739$_{-0.03}^{+0.03}$  \\
$\epsilon$ &    -1.005$_{-0.04}^{+0.03}$ &   -1.095$_{-0.02}^{+0.01}$ \\
$\sigma_g$ &    0.168$_{-0.07}^{+0.07}$ &    0.015$_{-0.03}^{+0.03}$  \\
$\zeta$ &    -0.014$_{-0.07}^{+0.07}$ & \\
    \hline
\end{tabular}
\end{table}
\label{tab:results}

@@

We have recalibrated the relation between short timescale brightness
fluctuations in the {\it Kepler} light curves of stars (flicker) with both
stellar density and surface gravity, whilst including parameters to describe
the intrinsic scatter in these relationships.
We included both an additive variance term {\it and} a term that allows for
some abscissa-dependent variance, presented in table \ref{tab:table}.
The additive terms, $\sigma_\rho$ and $ \sigma_g$ are both non-zero,
suggesting that there {\it is} an additional source of scatter in the
relations, not accounted for by the observational uncertainties alone.
This is either caused by intrinsic scatter in the physical relationship
between flicker and density and \logg, produced by some physical process that
is not accounted for in the model, or by an underestimation of the
observational uncertainties.
Interestingly, the abscissa-dependent variances, $\gamma$ and $\zeta$ are both
less than zero.
This indicates that the intrinsic scatter either decreases with increasing
flicker and decreasing stellar density and surface gravity, or that the
uncertainty underestimation is more severe for the giant stars.
Regardless of the cause of the additional scatter needed to describe the
relations between these parameters, the conclusions of this study are the same.
Firstly, naively modelling the relationship between flicker and surface
gravity or stellar density, or any other stochastic relationship
deterministically, will lead to an underprediction of the uncertainties in
dependent variable.
Secondly, we find marginal evidence for decreased scatter at higher surface
gravities and densities, i.e. the relations between flicker, surface gravity
and stellar density are valid for dwarfs and giants alike, despite the fact
that observational uncertainties tend to be larger for dwarfs.

This is a simple `fitting a line to data' exercise, however it continues the
discussion of probabilistic modelling that is an active topic within the
fields of exoplanet and stellar astronomy.
We used Hierarchical Bayesian Modelling (HBM) to constrain the intrinsic
scatter in the relationship between flicker, surface gravity and density.
We also included the effects of the non-negligable two-dimensional observational
uncertainties by marginalizing over latent variables using importance sampling.
% Although none of these methods, nor the data used here are new taken alone,
% we use the example of flicker to demonstrate the ideas and the importance of
% probabilistic, hierarchical inference.
Relationships between astronomical parameters are almost always
non-deterministic; an element of stochasticity effects the physical parameters
of stars so one can never perfectly predict $y$ given an observation of $x$.
We advocate a probabilistic approach in both the `fitting the model to data'
step, {\it and} when using an empirically calibrated model to predict
parameter values.
The fitting stage benefits because if the relationships between parameters are
falsely assumed to be deterministic, they will be skewed by data points with
unrepresentative uncertainties.
The prediction stage benefits from the stochastic treatment both because a
probability distribution is in many ways more representative of an observation
than a point estimate, and because posterior PDF samples can be used in
subsequent studies (provided the prior used during the fitting process is
described).
